A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-14 02-ceu.tex
Revisão Seção 2

\CEU is a synchronous reactive language in which programs evolve in a sequence
of discrete reactions to external events.
%
It is designed for control-intensive applications, supporting concurrent lines
of execution, known as \emph{trails}, and instantaneous broadcast communication
through events.
%
Computations within a reaction (such as expressions, assignments, and
system calls) are also instantaneous considering the synchronous
hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to execute its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment.

@@

In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate under the synchronous hypothesis, the compiler
environment should only provide access to~C operations that can be assumed to
be instantaneous, such as non-blocking~I/O and simple data structure accessors.%
\footnote {
The actual implementation of Céu supports a command-line option that accepts a
white list of library calls.
If a program tries to use a function not in this list, the compiler raises an
error.
}

@@

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A \CEU program with a ``subroutine''.},
  label={lst.sub},
]
event int* inc;         // declares subroutine "inc"
par/or do
    var int* p;
    every p in inc do   // implements "inc" through an event iterator
        *p = *p + 1;
    end
with
    var int v = 1;
    emit inc(&v);       // calls "inc"
    emit inc(&v);       // calls "inc"
    _assert(v==3);      // asserts result after the two returns
end
\end{lstlisting}
%\vskip-\baselineskip

@@

At first sight, event iteration such as in ``\code{every e do <...> end}'' seems
to be equivalent to ``\code{loop do await e; <...> end}''.
However, the \code{loop} variation would not compile because it does not
contain a path to an external input \code{await} (since \code{e} is an internal
event).
%
However, event iterators enforce other syntactic restrictions and cannot
contain \code{await} or \code{break} statements.
The absence of \code{break} guarantees that an iterator never terminates from
itself, essentially behaving as a safe blocking point in the program.
%
For this reason, the restriction that execution paths within loops must
contain at least one external \code{await} is extended to alternatively contain
an \code{every} statement.

@@

The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
sides in the composition need not be tweaked with synchronization primitives nor
state variables to affect each other.
%
In addition, abortion is \emph{immediate} in the sense that it executes
atomically in the current micro reaction.
%
Immediate orthogonal abortion is a distinctive feature of synchronous languages
and cannot be expressed effectively in traditional (asynchronous)
multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
    \item If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
          pointer represents a \textbf{local} resource (ln.~\ax:2) that
          requires finalization (ln.~\ax:7).
    \item If \CEU \textbf{receives} a pointer from a system call return
          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln.~11) before receiving an acknowledge from the
environment (ln.~9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln.~7).
%
In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln.~6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{%
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}

>>>>

\CEU~\cite{ceu.sensys13,ceu.tecs17} is a synchronous reactive language in which
programs evolve in a sequence of discrete reactions to external events.
%
It is designed for control-intensive applications, supporting concurrent lines
of execution, known as \emph{trails}, and instantaneous broadcast communication
through events.
%
Computations within a reaction (such as expressions, assignments, and
system calls) are also instantaneous considering the synchronous
hypothesis~\cite{rp.hypothesis}.
%
%\CEU provides an \code{await} statement which is the only that actually
%``consumes'' time.
%
\CEU provides an \code{await} statement that blocks the current running trail
allowing the program to execute its other trails; when all trails are blocked,
the reaction terminates and control returns to the environment.

@@

In \CEU, any identifier prefixed with an underscore (e.g., \code{\_led}) is
passed unchanged to the underlying~C compiler.
%
Therefore, access to~C is straightforward and syntactically traceable.
%
To ensure that programs operate under the synchronous hypothesis, the compiler
environment should only provide access to~C operations that can be assumed to
be instantaneous, such as non-blocking~I/O and simple data structure accessors.%
\footnote {
The actual implementation of Céu supports a command-line option that accepts a
white list of library calls.
If a program tries to use a function not in the list, the compiler raises an
error.
}

@@

\begin{lstlisting}[
  numbers=left,
  basicstyle=\ttfamily\footnotesize,
  float=h,
  caption={A ``subroutine'' that increments its argument.},
  label={lst.sub},
]
event int* inc;         // declares subroutine "inc"
par/or do
    var int* p;
    every p in inc do   // implements "inc" through an event iterator
        *p = *p + 1;
    end
with
    var int v = 1;
    emit inc(&v);       // calls "inc"
    emit inc(&v);       // calls "inc"
    _assert(v==3);      // asserts result after the two returns
end
\end{lstlisting}
%\vskip-\baselineskip

@@

At first sight, the constructions ``\code{every e do <...> end}'' and
``\code{loop do await e; <...> end}'' seem to be equivalent.
However, the \code{loop} variation would not compile since it does not contain
an external input \code{await} (\code{e} is an internal event).
%
The \code{every} variation compiles because event iterators have an additional
syntactic restriction that they cannot contain \code{break} or \code{await}
statements.
This restriction guarantees that an iterator never terminates from itself and,
thus, always awaits its identifying event, essentially behaving as a safe
blocking point in the program.
%
For this reason, the restriction that execution paths within loops must
contain at least one external \code{await} is extended to alternatively contain
an \code{every} statement.

@@

The \code{par/or} of \CEU is an orthogonal abortion mechanism because the two
sides in the composition need not be tweaked with synchronization primitives nor
state variables to affect each other.
%
In addition, abortion is \emph{immediate} in the sense that it executes
atomically inside the current micro reaction.
%
Immediate orthogonal abortion is a distinctive feature of synchronous languages
and cannot be expressed effectively in traditional (asynchronous)
multi-threaded languages~\cite{esterel.preemption,sync_async.threadsstop}.

@@

\CEU also enforces, at compile time, the use of \code{finalize} for system
calls that deal with pointers representing resources, as illustrated in the two
examples of Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
    \item If \CEU \textbf{passes} a pointer to a system call (ln.~\ax:5), the
          pointer represents a \textbf{local} resource (ln.~\ax:2) that
          requires finalization (ln.~\ax:7).
    \item If \CEU \textbf{receives} a pointer from a system call return
          (ln.~ \bx:4), the pointer represents an \textbf{external} resource
          (ln.~\bx:2) that requires finalization (ln.~\bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires
finalization clauses to accompany them.
%
In Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg}
(ln.~2) is an internal resource passed as a pointer to \code{\_send} (ln.~5),
which is an asynchronous call that transmits the buffer in the background.
If the block aborts (ln.~11) before receiving an acknowledge from the
environment (ln.~9), the local \code{msg} goes out of scope and the external
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln.~7).
%
In Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.
4) returns an external file resource as a pointer.
If the block aborts (ln.~12) during the \code{await A} (ln.~9), the file
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln.~6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{%
The compiler only forces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}
